Seg_sites_rarefaction

### 2023-10-17 scripts/seg_sites_rarefaction.Rmd
started work on paper

@@

# Generate data
```{r}
# Generate some example data:
## define parameters
allele_frequencies <- seq(.001, .1, by = .001) # allele frequencies
n1 <- 2000 # p1 gene copy number
n2 <- 200 # p2 gene copy number
k <- 30 # number to rarefact to
mcmcs <- 10000 # number of mcmc draws

## generate genotypes
g1 <- matrix(0, nrow = length(allele_frequencies), ncol = n1) # init
g2 <- matrix(0, nrow = length(allele_frequencies), ncol = n2)

m1 <- lapply(n1*allele_frequencies, function(x) sample(n1, x, FALSE))  # which are the minor
m2 <- lapply(n2*allele_frequencies, function(x) sample(n2, x, FALSE))

populate <- function(g, minors){g[minors] <- 1; return(g)} # function to replace minors
for(i in 1:length(allele_frequencies)){ # do for every locus
  g1[i,] <- populate(g1[i,], m1[[i]])
  g2[i,] <- populate(g2[i,], m2[[i]])
}

@@

# Via rarefaction approach
```{r}
counts <- function(g){
  return(data.frame(majhom = rowSums(g == 0), het = rowSums(g == 1), minhom = rowSums(g == 2)))
}
rarefact <- function(counts, k){
  return(1 - ((choose(counts[,1], k) + choose(counts[,3], k))/choose(rowSums(counts), k)))
}
varseg <- function(p) p*(1-p)

# taken from the binom package
confint_pb <- function(p, n, alpha){
  alpha <- rep(alpha, length = length(p))
  alpha2 <- 0.5 * alpha
  z <- qnorm(1 - alpha2)
  z2 <- z * z
  se <- sqrt(p * (1 - p)/n)
  lcl <- p - z * se
  ucl <- p + z * se

@@

# confint_sum <- function(nT, n, alpha){
#   Tbar <- nT/n
#   
#   if(n*Tbar == 0){return(0)}
#   
#   alpha_star <- zipfR::Rbeta((n*Tbar - 1)/n, n*Tbar, (n*(1 - Tbar)) + 1)
#   
#   if(alpha_star <= alpha & alpha <= 1){
#     top <- 1 - alpha
#     bottom <- n*(1 - alpha_star)
#     res <- Tbar - (top/bottom)
#   }
#   if(0 <= alpha & alpha <= alpha_star){
#     return(zipfR::Rbeta.inv(alpha, n*Tbar, n*(1 - Tbar) + 1))
#   }
#   
# }

@@

p1 <- sum(pseg1)/length(allele_frequencies)
p2 <- sum(pseg2)/length(allele_frequencies)
```

@@

```{r}
rarefact_mcmc <- function(g, k){
  samp <- sample(g, k, FALSE)
  seg <- ifelse(any(samp == 1), 1, # got hets
                ifelse(any(samp == 0) & any(samp == 2), 1, 0)) # got both homs
  return(seg)
}

# draw
seg1 <- seg2 <- matrix(0, length(allele_frequencies), mcmcs)
for(i in 1:mcmcs){
  for(j in 1:length(allele_frequencies)){
    seg1[j,i] <- rarefact_mcmc(g1[j,], k = k)
    seg2[j,i] <- rarefact_mcmc(g2[j,], k = k)
  }
}

@@

```{r}
# mcmcs <- 1000
# res <- d@geno.tables$gs
# res <- data.table::as.data.table(res)
# 
# matches <- which(d@facet.meta$facet == "pop")
# res <- res[matches,]
# 
# gs <- d@geno.tables$gs[matches,]
# gs <- data.table::as.data.table(gs)
# gs$.sum <- rowSums(gs) # sums for each row
# gs <- cbind(as.data.table(d@facet.meta[matches,]), gs)
# gs[,.g := min(.sum) - 1, by = .(facet, .snp.id)] # min across all levels
# 
# fwrite(res, "test_gs.txt", col.names = TRUE, sep = "\t", eol = "\n")
# fwrite(data.table(g = gs$.g), "test_g.txt", col.names = FALSE, sep = "\t", eol = "\n")
# fwrite(data.table(g = gs$subfacet), "test_subfacets.txt", col.names = FALSE, sep = "\t", eol = "\n")
# 
# 
# cmd <- paste0("perl seg_sites_rarefaction.pl test_gs.txt test_g.txt test_subfacets.txt ", mcmcs, " test_out.txt")
# system(cmd)
# 
# mcmc_res <- fread("test_out.txt")
# fc <- ncol(mcmc_res)
# mcmc_res <- mcmc_res[,-..fc]
# colnames(mcmc_res)[1] <- "subfacet"
# mcmc_res <- melt(mcmc_res, id.vars = "subfacet")
# 
# tr <- function(x){res <- t.test(x); return(list(seg_sites = res$estimate, lowerCI = res$conf.int[1], upperCI = res$conf.int[2]))}
# mcmc_summary <- mcmc_res[,tr(value), by = subfacet]

# results
```{r}
# individual loci
pseg <- rbind(data.frame(pop = paste0("n = ", n1/2), p = pseg1, snp = 1:length(allele_frequencies), af = allele_frequencies),
              data.frame(pop = paste0("n = ", n2/2), p = pseg2, snp = 1:length(allele_frequencies), af = allele_frequencies))

oseg <- rbind(data.frame(pop = paste0("n = ", n1/2), p = mpseg1[,1], snp = 1:length(allele_frequencies), af = allele_frequencies),
              data.frame(pop = paste0("n = ", n2/2), p = mpseg2[,1], snp = 1:length(allele_frequencies), af = allele_frequencies))

@@

ggplot(seg, aes(x = af, y = p, color = source)) + geom_point() + facet_wrap(~pop)

@@

ggplot(sum_tab, aes(x = pop, y = nseg)) + 
  geom_point(color = "red") +
  geom_errorbar(data = msum_tab, aes(ymax = upper, ymin = lower, x= pop), inherit.aes = FALSE) +
  # geom_point(data = msum_tab, color = "blue") +
  geom_jitter(data = all_mcmc_sums, height = 0, alpha = .005) +
  theme_bw() +
  scale_y_continuous(limits = c(0, length(allele_frequencies)))

@@

# estimated PI -- normal approx works well
pi1 <- qnorm(.975)*sqrt(sum(vseg1))*sqrt(1 + (1/length(allele_frequencies)))
pi2 <- qnorm(.975)*sqrt(sum(vseg2))*sqrt(1 + (1/length(allele_frequencies)))
pi <- data.frame(pop = paste0("n = ", n1/2), 
                 lower = sum(pseg1) - pi1,
                 upper = sum(pseg1) + pi1)
pi <- rbind(pi, data.frame(pop = paste0("n = ", n2/2), 
                 lower = sum(pseg2) - pi2,
                 upper = sum(pseg2) + pi2))

>>>>

```
# Define parameters
```{r}
# Generate some example data:
## define parameters
seed <- 231232
allele_frequencies <- seq(.001, .1, by = .001) # allele frequencies
n1 <- 2000 # p1 gene copy number
n2 <- 200 # p2 gene copy number
k <- 30 # number to rarefact to
mcmcs <- 10000 # number of mcmc draws
max_missing <- 30 # maximum percentage of missing data in a locus, from a uniform dist

## derived
set.seed(seed)
nal <- length(allele_frequencies)
miss_percent <- floor(runif(nal, 0, max_missing))/100

@@

# Generate data
```{r}
# generate genotypes
g1 <- matrix(0, nrow = nal, ncol = n1) # init
g2 <- matrix(0, nrow = nal, ncol = n2)

# add missing data
ms1 <- lapply(n1*miss_percent, function(x) sample(n1, x, FALSE))  # which are missing
ms2 <- lapply(n2*miss_percent, function(x) sample(n2, x, FALSE))

# determine minor allele placements
ma1 <- ma2 <- vector("list", length(ms1))
for(i in 1:nal){
  ma1[[i]] <- sample(n1*(1-miss_percent)[i], (n1*(1-miss_percent)*allele_frequencies)[i], FALSE)
  ma2[[i]] <- sample(n2*(1-miss_percent)[i], (n2*(1-miss_percent)*allele_frequencies)[i], FALSE)

}

# populate
## function to replace minors
populate <- function(g, missing, minors){
  g[missing] <- NA
  g[-missing][minors] <- 1
  return(g)
}
## do for every locus
for(i in 1:nal){ 
  g1[i,] <- populate(g1[i,], ms1[[i]], ma1[[i]])
  g2[i,] <- populate(g2[i,], ms2[[i]], ma2[[i]])
}

@@

# Via rarefaction approach
```{r}
counts <- function(g){
  return(data.frame(majhom = rowSums(g == 0, na.rm = TRUE), 
                    het = rowSums(g == 1, na.rm = TRUE), 
                    minhom = rowSums(g == 2, na.rm = TRUE)))
}
rarefact <- function(counts, k){
  return(1 - ((choose(counts[,1], k) + choose(counts[,3], k))/choose(rowSums(counts), k)))
}
varseg <- function(p) p*(1-p)

# taken from the binom package, asymptotic method
confint_pb <- function(p, n, alpha){
  alpha <- rep(alpha, length = length(p))
  alpha2 <- 0.5 * alpha
  z <- qnorm(1 - alpha2)
  z2 <- z * z
  se <- sqrt(p * (1 - p)/n)
  lcl <- p - z * se
  ucl <- p + z * se

@@

@@

p1 <- sum(pseg1)/nal
p2 <- sum(pseg2)/nal
```

@@

```{r}
rarefact_mcmc <- function(g, k){
  miss <- is.na(g)
  if(any(miss)){
    g <- g[-which(miss)]
  }

  samp <- sample(g, k, FALSE)
  seg <- ifelse(any(samp == 1), 1, # got hets
                ifelse(any(samp == 0) & any(samp == 2), 1, 0)) # got both homs
  return(seg)
}

# draw
seg1 <- seg2 <- matrix(0, nal, mcmcs)
for(i in 1:mcmcs){
  for(j in 1:nal){
    seg1[j,i] <- rarefact_mcmc(g1[j,], k = k)
    seg2[j,i] <- rarefact_mcmc(g2[j,], k = k)
  }
}

@@

# results
```{r}
# individual loci
pseg <- rbind(data.frame(pop = paste0("n = ", n1/2), p = pseg1, snp = 1:nal, af = allele_frequencies),
              data.frame(pop = paste0("n = ", n2/2), p = pseg2, snp = 1:nal, af = allele_frequencies))

oseg <- rbind(data.frame(pop = paste0("n = ", n1/2), p = mpseg1[,1], snp = 1:nal, af = allele_frequencies),
              data.frame(pop = paste0("n = ", n2/2), p = mpseg2[,1], snp = 1:nal, af = allele_frequencies))

@@

@@

ggplot(sum_tab, aes(x = pop, y = nseg)) + 
  geom_point(color = "red") +
  geom_errorbar(data = msum_tab, aes(ymax = upper, ymin = lower, x= pop), inherit.aes = FALSE) +
  geom_jitter(data = all_mcmc_sums, height = 0, alpha = .005) +
  theme_bw() +
  scale_y_continuous(limits = c(0, nal))

@@

# estimated PI -- normal approx works well
pi1 <- qnorm(.975)*sqrt(sum(vseg1))*sqrt(1 + (1/nal))
pi2 <- qnorm(.975)*sqrt(sum(vseg2))*sqrt(1 + (1/nal))
pi <- data.frame(pop = paste0("n = ", n1/2), 
                 lower = sum(pseg1) - pi1,
                 upper = sum(pseg1) + pi1)
pi <- rbind(pi, data.frame(pop = paste0("n = ", n2/2), 
                 lower = sum(pseg2) - pi2,
                 upper = sum(pseg2) + pi2))
